--- 
title: "Retrieve and standardise PJS-data"
output: 
  rmarkdown::html_vignette:
    css: "NVI_vignette_style.css"
    
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Retrieve and standardise PJS-data}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```
## Contents

-   [Overview](#overview)
-   [Set up the R environment](#set-up-the-r-environment)
-   [Retrive data from PJS](#retrive-data-from-pjs)
    -   [Make selection statement](#make-selection-statement)
    -   [Login to PJS](#login-to-pjs)
    -   [Read PJS-data](#read-pjs-data)
-   [Transform PJS-data](#transform-pjs-data)
    -   [Standardise PJS-data](#standardise-pjs-data)
    -   [Select variables from specified PJS levels](#select-variables-from-specified-pjs-levels)
    -   [Translate PJS codes into descriptive text](#translate-pjs-codes-into-descriptive-text)
-   [Further processing](#further-processing)

## Overview
The purpose of this vignette is to give guidelines for retrieving and standardising data from PJS. The vignette shows how functions in `NVIdb` can be used to facilitate this. Thereby, one should be able to make data sets from PJS with standard column names and standard classes for the variables to make it easier to share scripts for further management and analysis of PJS data. This vignette uses the wrapper function `retrieve_PJSdata` to show the most simple and efficient way for retrieving PJS data. The old vignette "Retrieve_and_standardise_PJS-data old-fashioned"

To run the scripts below, you will need R with the "NVIverse" packages `NVIbatch` and `NVIconfig` installed. You will also need a valid user account for PJS.

Be aware that all data, included PJS data, has pitfalls, peculiarities and mistakes. This vignette will make you aware of some of these. But remember; you should always make contact with the data owner, for example the scientist(s) responsible for the data, to ensure that you understand how the data has been generated and to ensure that the data can be used for the purpose you intend to use it for. Analysing data without sufficient knowledge on how the data has been generated may give strange results.

## Set up the R environment
In this part you want to set up the R environment for all scripts that are called (sourced) by the main script. This includes attaching all packages needed and setting global variables and functions.

### Attach R packages
In base R, one use `library` or `require` to attach packages. Of these, `library` is preferred as `require` will not produce an error message if a package is not attached. The packages have to be installed before using `library` or `require`.
```{r, include = TRUE}
# SET UP THE R ENVIRONMENT ----
# # Attach R packages (using base R) ----
library(dplyr)
library(RODBC)
library(NVIdb)
```

An alternative is to use `use_pkg` and `use_NVIverse` from the package `NVIbatch`. These functions accept a vector with package names and will attach the packages if installed, and install the package if not installed. Use `use_pkg` for packages at "Cran" and `use_NVIverse` for packages within the "NVIverse". If the packages need to be installed, you also need writing access to the directory for R packages. 
```{r, include = TRUE}
# SET UP THE R ENVIRONMENT ----
# # Attach R packages (using NVIbatch) ----
library(NVIbatch)
use_pkg(pkg = c("dplyr", "RODBC"))
use_NVIverse(pkg = c("NVIdb"))
```

### Set global variables and functions
The global variables should be defined at the beginning of the script to make them easy to find and eventually edit. In this example there are one global variable: today's date. 

Likewise should global functions used by the script be defined at the beginning of the script. These are functions needed for the script that are not included in any package, either because they are very specific for this script or that they are not yet mature for being included into a package. If functions are used in several scripts, it is recommended to generalise the functions and move them to packages to facilitate maintenance, updating and error testing.

Instead of including the code for defining the global variables and the functions in the main script, the code may be written in separate scripts (R files) that are sourced from the main script.
```{r, include = TRUE}
# # Global variables ----
# ## Paths and file names
# None in this example

# ## variables
# Today's date in the format yyyymmdd for use in file names.
today <- format(Sys.Date(), "%Y%m%d")

# ## Functions
# None in this example
```

### Import support registers
Very often, you will combine PJS-data with information found in other registers. In this example we use tables to translate PJS codes to descriptive text. Currently, when using `NVIdb`, such registers must be loaded/imported before they can be used. This may change when this information is easily accessible through databases.
```{r, include = TRUE}
# # Import support registers ----
# Translation tables for PJS codes
PJS_codes_2_text <- read_PJS_codes_2_text()
```

## Retrive data from PJS
We retrieve data from PJS through views in the data base journal_rapp. There are views available for different purposes. If we need all data for a disease, we usually use the view "v2_sak_m_res" that combines information from all levels from sak to result and in addition includes konklusjon and eier_lokalitet. 

### Make selection statement
In this example we select all saker with resultater and konklusjoner related to surveillance for paratuberculosis in different species for a single year. To achieve this we select

- all saker with hensikt for paratuberculosis surveillance. 

There are four main hensikter, each for one species (group of species). For three of them there are sub-hensikter. To retreive data for sub-hensikter when using `build_query_hensikt`, you can add % at the end of the code instead of listing all the sub-hensikter.
```{r, include = TRUE}
# RETRIEVE DATA FROM PJS ----

# Set selection parameters
purpose <- "Paratuberkulose"
aar <- c(2020)
hensikt2select <- c("0200106%", "0200119", "0200122%", "0200125%")
```

We can use build_query_hensikt to generate the sql syntax for selecting observations from the views v2_sak_res_eier and sakskonklusjon in journal_rapp. The select statement retrieves all observations from one (or more) years AND the hensikter with or without sub-hensikter that you give in the input statement. The select statement is used as input to the query sent to the database.
```{r, include = TRUE}
select_statement <- NVIdb::build_query_hensikt(year = aar,
                                               hensikt = hensikt2select)
```


### Login to PJS
To retrieve the PJS data, you log in to the database journal_rapp to established a data base connection, query the database and close the database connection. You can use `login_PJS` (equals `login("PJS")`) for the log in. You will then enter the username and password, but the other specifications of the data base connection are not necessary to input. If you often login to PJS, you may store the credentials (username and password) in your user profile (at the current PC) with `set_credentials_PJS`. These will be automatically found by `login`. Must be repeated when password to PJS changes or if you use another PC. To avoid being asked for credentials when running batch scripts, store the credentials with `set_credentials_PJS` and use `login_by_credentials_PJS`.
```{r, include = TRUE}
# READ DATA FROM PJS ----
journal_rapp <- login_PJS()
```

### Read PJS data
The query can be sent PJS by `RODBC::sqlQuery`. In this example we only select observations using the view "v2_sak_res_eier" and not from the view "sakskonklusjon". The argument `as.is =TRUE` ensures that leading zeros are kept in character variables with only numbers.

After the wanted queries have been run, you should tidy up by closing the data base connection. 
```{r, include = TRUE}
PJSrawdata <- sqlQuery(journal_rapp,
                       select_statement["selection_v2_sak_m_res"],
                       as.is = TRUE,
                       stringsAsFactors = FALSE)


PJSsakskonklusjon <- sqlQuery(journal_rapp,
                              select_statement["selection_sakskonklusjon"],
                              as.is = TRUE,
                              stringsAsFactors = FALSE)

odbcClose(journal_rapp)
```

## Transform PJS-data

### Standardise PJS-data 
To facilitate writing script processing PJS data, we have developed column name standards and use functions to automatically set these names. In addition, correct class for numeric and date variables are ensured. The function `standardize_PJSdata` performs the following transformations:

- The column names are standardized using `standardize_columns`. 
- Numeric variables are transformed from character to numeric. 
- Date variables are transformed to date class. 
- Character variables are trimmed for leading and trailing spaces. 
- The variables "saksnr" and, if possible, "fagnr" are generated. 
- The unnecessary columns "konkl_provenr" and "vet_distriktnr" are removed. 
- Test data, i.e. saker with "ansvarlig_seksjon" in c("14", "99") are deleted. 

Thereafter, samples that are irrelevant for reporting Norwegian data, i.e. quality assessment and samples from abroad, can be excluded. Be aware that only samples registered with owner/location type = country are excluded when excluding samples from abroad. If the owner/location type is an organisation, firm, laboratory or person with address in another country, these are not excluded. 
```{r, include = TRUE}
# # STANDARDIZE DATA ----
# ## Standard cleaning of data
PJSdata <- standardize_PJSdata(PJSdata = PJSrawdata)
sakskonklusjon <- standardize_PJSdata(PJSdata = PJSsakskonklusjon)

# # Exclude ring trials, quality assurance and samples from abroad
PJSdata <- exclude_from_PJSdata(PJSdata = PJSdata, abroad = "exclude", quality = "exclude")
sakskonklusjon <- exclude_from_PJSdata(PJSdata = sakskonklusjon, abroad = "exclude", quality = "exclude")
```

### Select variables from specified PJS levels
Several views retrieve data from several of the PJS levels as in this example. By selecting unique rows from specific PJS levels, the number of rows in the data may be reduce considerably. You may select variables from specified PJS levels by using the function `choose_PJS_levels`. The function accepts the levels as input and you do not have to specify all the variables within the levels. The function will only work as intended if the column names have been standardised using `standardize_PJS`. In this example we remove the columns "konklnr" and "konkl_type". Thereby, we remove duplicate konklusjoner when `unique_rows = TRUE`. See `choose_PJS_levels` for further options. 
```{r, include = TRUE}
# # KEEP VARIABLES IN SAK, PROVE AND KONKLUSJON LEVEL ----
PJSdata <- choose_PJS_levels(data = PJSdata,
                             levels = c("sak", "prove", "konklusjon"),
                             keep_col = NULL,
                             remove_col = c("konklnr", "konkl_type"),
                             unique_rows = TRUE)
```

### Translate PJS codes into descriptive text
To translate PJS codes into descriptive text you may use the function `add_PJS_code_description`. The function needs input of the data, the translation table, the name  of the columns with the codes, the name of the new columns with the descriptive text and the type of the code that should be translate. There are additional arguments to position the new columns etc., see `add_PJS_code_description`. When the column names have been standardised as done above, you can simplify the input arguments code_column and new_column as shown in the example below. See `add_PJS_code_description` for how to do it with non-standard names. 

The code below is dependent on the translation table already have been imported in a previous step. In this case it was imported in the section [Import support registers](#import-support-registers).
```{r, include = TRUE}
# # TRANSLATE CODES INTO DESCRIPTIVE TEXT  ----

# Translate PJS-codes to code descriptions, simplified version
PJSdata <- add_PJS_code_description(data = PJSdata,
                                    translation_table = PJS_codes_2_text,
                                    PJS_variable_type = "auto",
                                    code_colname = c("hensiktkode", "artkode", "driftsformkode", "provetypekode",
                                                     "provematerialekode", "kjonn", "konkl_kjennelsekode", "konkl_analyttkode"),
                                    new_column = "auto",
                                    position = "right",
                                    overwrite = FALSE)
```

## Further processing
The resulting data frame PJSdata includes selected observations from PJS from the levels: sak, prove and konklusjon. These is available for performing further analysis, descriptive statistics etc. In this example we save the data for further use. 
```{r, include = TRUE}
# SAVE DATA ----
saveRDS(PJSdata, file = file.path(tempdir(), paste0(purpose, ".rds")))
```


